Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-07-05 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
revised manuscript

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
%\RequirePackage{fix-cm}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}
\usepackage{nicefrac}

@@

\institute{Gabriel Riutort-Mayol \at
              Department of Cartographic Engineering, Geodesy, and Photogrammetry, Universitat Polit\`ecnica de Val\`encia, Spain\\
              \email{gabriuma@gmail.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Excellence Cluster for Simulation Technology, University of Stuttgart, Germany \\
              \emph{Most of work was done while at Dept. of Computer Science at Aalto University}
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland
}

@@

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity of the function to be estimated, which is in turn characterized by the length-scale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the squared exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the smoothness or wigglyness of the covariance function.

@@

\begin{figure*} %[!t]
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 3mm 0mm, clip]{fig1_Sigma_J.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig1_Cov_J.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), posterior standard deviations (center), and covariance functions (right) of both the exact GP model (dashed red line) and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure*}

\begin{figure*}
\centering
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 0mm 5mm 3mm 0mm, clip]{fig2_Sigma_L.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 1mm 5mm 3mm 0mm, clip]{fig2_Cov_L.pdf}}
\vspace{-1mm}
\caption{Mean posterior predictive functions (left), posterior standard deviations (center), and covariance functions (right) of both the exact GP model (dashed red line) and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure*}

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of length-scale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the length-scale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For a sufficiently large fixed value of $c$, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig2_Post_L} shows that $c$ affects the approximation of the mean mainly near the boundaries, while the approximation of the standard deviation is affected across the whole domain. The approximation error tends to be bigger for the standard deviation than for the mean.

\begin{figure*}
\centering
\begin{tabular}{ c c c }
\arrayrulecolor{darkgray}\hline
c = 1.05 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Post_part1.pdf}
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Sigma_part1.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{fig3_Cov_part1.pdf} & 
\multirow{43}{*}{ \includegraphics[scale=0.35, trim = 28mm 30mm 100mm 30mm, clip]{fig3_legend.pdf}}\\ 
\arrayrulecolor{lightgray}\cline{1-2}
c = 1.1 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Post_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Sigma_part2.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{fig3_Cov_part2.pdf} &\\
\cline{1-2}
c = 1.2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Post_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Sigma_part3.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{fig3_Cov_part3.pdf} &\\
\cline{1-2}
c = 1.5 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Post_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Sigma_part4.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{fig3_Cov_part4.pdf} & \\
\cline{1-2}
c = 2 &
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Post_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 0mm 14mm, clip]{fig3_Sigma_part5.pdf} 
\includegraphics[scale=0.215, trim = 0mm 14mm 5mm 14mm, clip]{fig3_Cov_part5.pdf} & \\
\cline{1-2}
c = 2.5 &
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{fig3_Post_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 0mm 14mm, clip]{fig3_Sigma_part6.pdf}
\includegraphics[scale=0.215, trim = 0mm 4mm 5mm 14mm, clip]{fig3_Cov_part6.pdf} &\\
\arrayrulecolor{darkgray}\hline
\end{tabular}
\caption{Posterior mean predictive functions (left), posterior standard deviations (center) and  covariance functions (right) of both the exact GP model and the HSGP model for different number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig3_Post_part1}
\end{figure*}

\begin{figure*} %[!t]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig4_MSE_vs_c.pdf}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
  \label{fig4_MSE_vs_J}
\end{figure*}

\begin{figure*} %[!h]
\centering
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_lscale_vs_J.pdf}}
\subfigure{\includegraphics[scale=0.38, trim = 0mm 8mm 0mm 0mm, clip]{fig5_magnitud_vs_J.pdf}}
\caption{Estimated length-scale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig5_lscale_vs_J}
\end{figure*}

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig3_Post_part1} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and~$c$. Figure~\ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure~\ref{fig5_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure~\ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$ could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:
%
\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

@@

When considering squared exponential and Mat{\'e}rn covariance functions, we can show with simple algebra that when $c$ is larger than the minimal value recommendation, the number of $m$ first terms needed to explain almost 100\% of the variation has a near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$, and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, as a smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$, the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$, if $\ell$ or $c$ are changed.

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig6_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig6_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $\frac{\ell}{S}$ (lengthscale $\ell$ relative to the half-range $S$ of the input domain). Figure \ref{fig6_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2), and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with length-scale $\frac{\ell}{S}$ (From now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$) and given a boundary factor $c$, Figure \ref{fig6_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which an accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig6_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near linear proportionality between $m$, $\ell$ and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

Figure~\ref{fig6_relationships} and previous equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure~\ref{fig6_relationships} or equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

As commented in Section~\ref{sec_method_periodic}, in Appendix~\ref{sec_periodic} we present an approximate linear representation of a periodic squared exponential covariance function. In Appendix~\ref{sec_periodic}, we also analyze the accuracy of this linear representation and derive the minimum number of terms $m$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition. This relationship between $m$ and $\ell$ for a periodic squared exponential covariance function is gathered in Figure~\ref{figB1_m_lscale_periodic} and the numerical equation was estimated in eq.~\eqref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m} .
\end{align*}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
%%!PS-Adobe-3.0 EPSF-3.0
%%%BoundingBox: 19 19 221 221
%%%CreationDate: Mon Sep 29 1997
%%%Creator: programmed by hand (JK)
%%%EndComments
%gsave
%newpath
%  20 20 moveto
%  20 220 lineto
%  220 220 lineto
%  220 20 lineto
%closepath
%2 setlinewidth
%gsave
%  .4 setgray fill
%grestore
%stroke
%grestore
%\end{filecontents*}
%
%\RequirePackage{fix-cm}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{blindtext}
%\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{times}
\usepackage{microtype}
\usepackage{lscape}
\usepackage{epstopdf}
\usepackage{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{float}
\usepackage{soul}
\usepackage[dvipsnames]{xcolor}
\usepackage{amssymb}
\usepackage{amsmath}	
\usepackage{mathtools}
%\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}
\usepackage{pbox}
\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{caption}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,breaklinks=true]{hyperref}
\usepackage{nicefrac}

@@

\institute{Gabriel Riutort-Mayol \at
              Foundation for the Promotion of Health and Biomedical Research of Valencia Region (FISABIO), Spain\\
              \emph{Corresponding Author} \\
              \email{gabriuma@gmail.com; riutort\_gab@gva.es}       %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Paul-Christian Bürkner \at
              Excellence Cluster for Simulation Technology, University of Stuttgart, Germany \\
              \emph{Most of work was done while at Dept. of Computer Science at Aalto University}
           \and
           Michael R.\ Andersen \at
              Department of Applied Mathematics and Computer Science, Technical University of Denmark, Denmark
           \and
           Arno Solin \at
              Department of Computer Science, Aalto University, Finland
           \and
           Aki Vehtari \at
              Department of Computer Science, Aalto University, Finland\\
              \email{aki.vehtari@aalto.fi}
}

@@

The accuracy and speed of the HSGP model depends on several interrelated factors, most notably on the number of basis functions and on the boundary condition of the Laplace eigenfunctions. Furthermore, appropriate values for these factors will depend on the degree of non-linearity ({\color{blue}wigglyness/smoothness}) of the function to be estimated, which is in turn characterized by the length-scale of the covariance function. In this section, we analyze the effects of the number of basis functions and the boundary condition on the approximation accuracy. We present recommendations on how they should be chosen and diagnostics to check the accuracy of the obtained approximation.

Ultimately, these recommendations are based on the relationships among the number of basis functions $m$, the boundary condition $L$, and the length-scale $\ell$, which depend on the particular choice of the kernel function. In this work we investigate these relationships for the squared exponential and the Mat{\'e}rn ($\nu=3/2$ and $\nu=5/2$) covariance functions in the present section, and for the periodic squared exponential covariance function in Appendix~\ref{sec_periodic}. For other kernels, the relationships will be slightly different depending on the {\color{blue}smoothness or wigglyness of the non-linear effects generated from the covariance function}.

@@

\begin{figure}
%\centering
\includegraphics[width=0.485\textwidth, trim = 2mm 0mm 0mm 0mm, clip]{fig1_Post_J_L.png}
\vspace{-1mm}
\caption{Mean posterior predictive functions (top), posterior standard deviations (center), and covariance functions (bottom) of both the exact GP model (red line) and the HSGP model for 1) different number of basis functions $m$, with the boundary factor fixed to a large enough value (left column) and 2) different values of the boundary factor $c$, with a large enough fixed number of basis functions (right column).}
  \label{fig1_Post_J_L}
\end{figure}

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig2_Post.png}}
\vspace{-1mm}
\caption{Posterior mean predictive functions (top), posterior standard deviations (center) and  covariance functions (bottom) of both the exact GP model and the HSGP model for different number of basis functions $m$ and for different values of the boundary factor $c$ (columns).}
  \label{fig2_Post}
\end{figure*}

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of length-scale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the length-scale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J_L} and \ref{fig1_Post_J_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For a sufficiently large fixed value of $c$, Figure \ref{fig1_Post_J_L} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig1_Post_J_L} shows that $c$ affects the approximation of the mean mainly near the boundaries, while the approximation of the standard deviation is affected across the whole domain. The approximation error tends to be bigger for the standard deviation than for the mean.

Next, we analyze how the interaction between $m$ and $c$ affects the quality of the approximation. The length-scale and marginal variance of the covariance function will no longer be fixed but instead we compute the joint posterior distribution of the function values and the hyperparameters using the dynamic HMC algorithm implemented in Stan \citep{StanTeam:2021} for both the exact GP and the HSGP models. Figure~\ref{fig2_Post} shows the posterior predictive mean and standard deviation of the function as well as the covariance function obtained after fitting the model for varying $m$ and~$c$. Figure~\ref{fig3_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models computed against the exact GP model. Figure~\ref{fig4_lscale_vs_J} shows the estimated length-scale and marginal variance for the exact GP model and the HSGP models. Looking at the RMSEs in Figure~\ref{fig3_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computation time for this example would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the less conservative choice of $m = 10$ and $c = 1.5$ could also produce a sufficiently accurately approximation depending on the application. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure~\ref{fig2_Post}. From these results, some general conclusions may be drawn:
%
\begin{itemize}
\item As $c$ increases, $m$ has to increase as well (and vice versa). This is consistent with the expression for the eigenvalues in eq.~\eqref{eq_eigenvalue}, where $L$ appears in the denominator.
\item There exists a minimum $c$ below which an accurate approximation will never be achieved regardless of the number of basis functions $m$.
\end{itemize}

\begin{figure} %[!t]
\centering
\subfigure{\includegraphics[width=0.49\textwidth, trim = 2mm 0mm 3mm 0mm, clip]{fig3_MSE_vs_J.png}}
\caption{Root mean square error (RMSE) of the proposed HSGP models computed against the exact GP model. RMSE versus the number of basis functions $m$ and for different values of the boundary factor $c$ (left). RMSE versus the boundary factor $c$ and for different values of the number of basis functions $m$ (right). }
\label{fig3_MSE_vs_J}
\end{figure}

\begin{figure} %[!t]
\centering
\subfigure{\includegraphics[width=0.49\textwidth, trim = 2mm 0mm 2mm 0mm, clip]{fig4_lscale_vs_J.png}}
\caption{Estimated length-scale (left) and marginal variance (right) parameters of both exact GP and HSGP models, plotted versus the number of basis functions $m$ and for different values of the boundary factor $c$.}
\label{fig4_lscale_vs_J}
\end{figure}

@@

When considering squared exponential and Mat{\'e}rn covariance functions, we can show with simple algebra that when $c$ is larger than the minimal value recommendation, the number of $m$ first terms needed to explain almost 100\% of the variation has a near linear relationship with $\ell$ and $c$. With decreasing $\ell$, the $m$ should grow near linearly with $1/\ell$, and with increasing $c$, the $m$ should grow near linearly with $c$.
%
This is natural as with decreasing $\ell$, more higher frequency basis functions are needed. With increasing $c$, as a smaller range of the basis functions are used in the approximation, the expected number of zero up-crossings goes down linearly with $c$, and thus more higher frequency basis functions are needed to compensate this.
%
When $c$ is below our recommendations given $\ell$ ({\color{blue}recommendations that we give later throughout the paper and specifically in the next Sections \ref{subsec_empiric_relations} and \ref{sec_num_equations}}), the effect of $\ell$ and $c$ to the recommended $m$ is more non-linear, but as long as we stay in the recommended range the linearity assumption is useful thumb rule for how to change $m$, if $\ell$ or $c$ are changed.

@@

\begin{figure*}
\centering
\subfigure{\includegraphics[width=\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig5_relationships.png}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ \hspace{0.1mm} ($c = \frac{L}{S}$) \hspace{0.1mm} and the length-scale normalized by the half-range of the data ($\frac{\ell}{S}$), for the squared exponential, Mat\'ern ($\nu$=3/2) and Mat\'ern ($\nu$=5/2) covariance functions.}
  \label{fig5_relationships}
\end{figure*}

Empirical simulations are carried out to analyze the relationships between $m$, $c$ and $\frac{\ell}{S}$ (lengthscale $\ell$ relative to the half-range $S$ of the input domain). Figure \ref{fig5_relationships} depicts how these three factors interact and affect the accuracy of the HSGP approximation for a GP with squared exponential, Mat\'ern ($\nu$=3/2), and Mat\'ern ($\nu$=5/2) covariance functions and a single input dimension. More precisely, for a given GP model (with a particular covariance function) with {\color{blue}normalized} length-scale $\frac{\ell}{S}$ (From now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$) and given a boundary factor $c$, Figure \ref{fig5_relationships} shows the minimum number of basis functions $m$ required to obtain an accurate approximation in the sense of satisfying eq. (\ref{eq_diff_covs}). We considered an approximation to be a sufficiently accurate when the total variation difference between the approximate and exact covariance functions, $\varepsilon$ in eq. (\ref{eq_diff_covs}), is below 1$\%$ of the total area under the curve of the exact covariance function $k$ such that
%
\begin{equation} \label{eq:rel_total_distance}
 \int | k(\tau) - \tilde{k}_m(\tau)|  \,\mathrm{d}\tau  < 0.01 \int k(\tau) \,\mathrm{d}\tau,
\end{equation}
%
where $\tilde{k}_m$ is the approximate covariance function with $m$ basis functions. Alternatively, these figures can be understood as providing the minimum $c$ that we should use for given $\ell$ and $m$. Of course, we may also read it as providing the minimum $\ell$ that can be approximated with high accuracy given $m$ and $c$. We obtain the following main conclusions:
%
\begin{itemize}
\item As $\ell$ increases, $m$ required for an accurate approximation decreases. ({\color{blue}Notice that the larger $\ell$ the smoother the functions generated from a covariance function, and vice versa; see covariance functions equations in Section \ref{ch4_sec_cov}.})
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve an accurate approximation.
\item For a given $\ell$ there exist a minimum $c$ under which an accurate approximation is never going to be achieved regardless of $m$. This fact can be seen in Figure~\ref{fig5_relationships} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$). As $\ell$ increases, the minimum valid $c$ also increases.
\end{itemize}

@@

As explained in Section~\ref{subsec_theoretic_linearity}, when $c$ is large enough, there is a near linear proportionality between $m$, $\ell$ and $c$. To obtain practical numerical functions that can be used to guide the selection of these parameters, we have empirically ({\color{blue}checked this linear relationship and} derived the practically useful constant terms. We require a lower bound for $c$ of $c \geq 1.2$ such that the equations below are precise enough for practical application.

@@

{\color{blue}Note that in the above equations we have made explicit the use of the normalized length-scale by the half-range of the input space ($\frac{\ell}{S}$) (From now to later on we will refer to $\frac{\ell}{S}$ by just $\ell$). By using the normalized length-scale, the equations, figures, and conclusions made throughout the paper can easily be referred to any specific scale of the input space.}

Figure~\ref{fig5_relationships} and previous equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) were obtained for a GP with a unidimensional covariance function, which result in a surfaces depending on three variables, $m$, $c$ and $\ell$. Equivalent results for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wigglyness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figure~\ref{fig5_relationships} or equations~(\ref{eq_m_l_QE})-(\ref{eq_c_vs_l_mat52}) to check the accuracy by analyzing individually the different dimensions of a multidimensional GP model.

@@

As commented in Section~\ref{sec_method_periodic}, in Appendix~\ref{sec_periodic} we present an approximate linear representation of a periodic squared exponential covariance function. In Appendix~\ref{sec_periodic}, we also analyze the accuracy of this linear representation and derive the minimum number of terms $m$ in the approximation required to achieve a close approximation to the exact periodic squared exponential kernel as a function of the length-scale $\ell$ of the kernel. Since this is a series expansion of sinusoidal functions, the approximation does not depend on any boundary condition{\color{blue}, nor is there a need for a normalized length scale, since the length-scale refers to the period of a sinusoidal function}. This relationship between $m$ and $\ell$ for a periodic squared exponential covariance function is gathered in Figure~\ref{figB1_m_lscale_periodic} and the numerical equation was estimated in eq.~\eqref{eq_J_l_periodic} which is depicted next:
%
\begin{align*}
&m \geq \frac{3.72}{\ell} \;\; \Leftrightarrow \;\; \ell \geq \frac{3.72}{m} .
\end{align*}

@@

({\color{blue}In Section \ref{sec_requirements_diagnosis} readers can find a brief remark on the iterations requirements in the diagnosis procedure.})
